Learning pleiotropic signatures to construct a causal map of the human phenome

### 2017-08-02 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Learning pleiotropic signatures to construct a causal map of the human
phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{29 July 2017}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  MRC Integrative Epidemiology Unit (IEU) at the University of Bristol,
  School of Social and Community Medicine, Bristol, UK
\item
  Intelligent Systems Laboratory, Department of Computer Science,
  University of Bristol
\end{enumerate}

@@

A major application for genome-wide association studies (GWAS) has been
the emerging field of causal inference using Mendelian randomization
(MR), where the causal effect between a pair of traits can be estimated
using only summary level data. MR depends on SNPs exhibiting vertical
pleiotropy, where the SNP influences an outcome phenotype only through
an exposure phenotype. Issues arise when this assumption is violated due
to SNPs exhibiting horizontal pleiotropy, and many methods have been
developed in an attempt to address this. Here we show that the
mechanisms that underlie horizontal pleiotropy are numerous, and that
instrument selection will be increasingly liable to selecting invalid
instruments as GWAS sample sizes continue to grow. We have developed a
mixture of experts machine learning approach (MR-MoE 1.0) that improves
on both power and false discovery rates over unselective use of any
existing methods. Using the approach, we systematically estimated the
causal effects amongst 2407 phenotypes, generating a working draft of
the causal map of the human phenome comprising the evaluation of
causality between over a million pairs of traits. Almost 90\% of causal
estimates indicated some level of horizontal pleiotropy. The causal
estimates are organised into a publicly available graph database
(\url{http://eve.mrbase.org}).

@@

Mendelian randomization (MR) (1,2) exploits genetic pleiotropy to infer
the causal relationships between phenotypes. Suppose that one trait (the
exposure) causally influences another (the outcome). If a SNP influences
the outcome through the exposure then the SNP is exhibiting vertical
pleiotropy. Such a genetic variant is considered to be an instrumental
variable for the if it only influences the outcome through the exposure
(the exclusion restriction). It can be exploited to mimic a randomised
controlled trial, enabling a causal estimate to be made by comparing the
outcome phenotypes between those individuals that have the
exposure-increasing allele against those who do not. Multiple
independent genetic variants for a particular exposure can be used
jointly to improve causal inference, because a) each variant represents
an independent natural experiment, and an overall causal estimate can be
obtained by meta-analysing the single estimates from each instrument;
and b) potential bias arising from the exclusion restriction principal
can be detected or corrected by evaluating the consistency of effects
across instruments (3--8).

Genome-wide association studies (GWAS) have identified genetic
instrumental variables for thousands of phenotypes (9). Recent
developments in MR have enabled knowledge of instrumental variables to
be applied using only summary level data (known as two-sample MR, 2SMR)
(10). Here, in order to infer the causal effect of an exposure on an
outcome all that is required is an estimate of the genetic effects of
the instrumenting SNP on the exposure, and the corresponding estimate of
the effect on the outcome. This has three major advantages. First, GWAS
summary data are non-disclosive and often publicly available. Second,
causal inference can be made between phenotypes even if they have not
been measured in the same samples, limiting possible causal estimates
only by the availability of GWAS summary data for the traits in question
(11). Third, we can view the causal inference problem through the simple
and widely understood prism of a meta-analysis.

Problems with obtaining unbiased causal effects can arise, however, if
the genetic instruments exhibit horizontal pleiotropy (HP), where they
influence the outcome through a pathway other than the exposure (2). The
extent of this phenomenon is not to be understated, and many methods
have been developed that attempt to reliably obtain unbiased causal
estimates under specific models of HP (3--6,8,12). It is considered best
practice to report estimates from all available methods as sensitivity
analyses when presenting causal estimates because different pleiotropic
models can then be evaluated. However it is also desirable to
consolidate across methods under some circumstances. First, it could be
of interest to make causal effect estimates for thousands of traits, in
which case a critical evaluation of each causal estimate of interest may
not be possible or convenient. Second, though the IVW approach is most
statistically powerful under no HP, it can have high false negative or
low true positive rates in the presence of HP compared to other methods.
Pleiotropy has been hypothesised to be universal (13), though the degree
to which this may be the case is contested (14,15), hence defaulting to
the IVW method in the first instance and using other methods as
sensitivity analyses may not be appropriate. Third, if different methods
disagree it is not possible to know which is correct because the true
nature of HP exhibited by the instruments is not known. Fourth, the
available methods do not cover all possible models of HP, and therefore
an automated method for instrument selection may be necessary.

In this paper we introduce two innovations towards improving the
reliability of MR estimates. First, we introduce an approach to discard
genetic variants that are likely to be invalid. Second, we hypothesised
that characteristics of the summary data could indicate which method
would be most reliable, and we introduce new machine learning approaches
that attempt to automate both instrument and method selection. Using
curated GWAS summary data for thousands of phenotypes (11), we use these
new methods to construct a graph of millions of causal estimates.
Motivated by the recent avalanche (16) of 2SMR publications, often with
repeating (17,18) or contradicting (19,20) conclusions despite using the
same data, we developed a graph database to represent these estimates in
a consistent manner. We consider this to be a `working draft' of the
causal map of the human phenome, but raise caution throughout that its
interpretation is far from straightforward, and that future corrections
and refinements will inevitably follow as data grows and 2SMR methods
evolve.

@@

The use of summary data in two-sample MR is described in detail
elsewhere (11). A brief outline of the procedure is as follows. First,
genetic instruments for the exposure trait need to be identified - those
SNPs with \(p < 10^{-8}\) are retained in order to ensure that the first
assumption of MR (that the instrument associates with the exposure) is
generally satisfied. We collect their effect sizes, standard errors and
effect alleles for the association with the exposure trait. These can be
obtained manually from complete summary data or from curated lists of
significant GWAS associations. Next, the effects of those SNPs on the
outcome need to be obtained, typically necessitating complete summary
data because these SNPs will generally have not have reached genome-wide
significance for other traits (and therefore be present in curated
catalogues).

We use the term \emph{summary-set} to refer to the minimum data required
to perform 2SMR analysis - four columns comprising the SNP-exposure and
SNP-outcome effects and standard errors, and rows corresponding to each
SNP that is used as an instrument for the exposure.

At its most simple implementation, the regression of the SNP-exposure
effect sizes against the SNP-outcome effect sizes, with greater weight
afforded to those SNPs with smaller SNP-outcome standard errors,
provides the estimate of the causal effect of the exposure on the
outcome. This is known as the inverse-variance weighted (IVW) method.

Given summary data for a large number of traits, it is straightforward
to exhaustively analyse the causal relationships of every trait against
every other trait for which there are sufficient summary data available.
Supplementary table 1 provides a list of all traits that have available
GWAS summary data, indicating if they have complete summary data (in
which case they can be used as both exposure traits and outcome traits),
or if they only have significant associations (in which they can only be
used as exposure traits).

@@

In this paper we consider three main classes of MR estimation. Full
details for each approach have been described previously. A summary of
the methods is given in Supplementary table 2.

\textbf{Mean-based methods:} Here we consider four nested models (5).
The inverse variance weighted (IVW) fixed effects meta-analysis approach
assumes that variants exhibit no HP. IVW random effects meta-analysis
relaxes the HP assumption, allowing it to be present but balanced - such
that it only leads to increased heterogeneity around the regression and
not introducing bias. Fixed effects Egger regression (3) relaxes the HP
assumption further by allowing a non-zero intercept which essentially
allows horizontal pleiotropy to be directional, where it systematically
influences the outcome in a specific direction. Random effects Egger
regression allows heterogeneity around the directional HP (5), as long
as the HP effects are not correlated with the SNP-exposure effects (also
known as the INSIDE assumption)(3).

The Rucker framework (21), adapted to MR (5) uses estimates of
heterogeneity in the IVW and Egger frameworks to navigate between these
nested models. A jackknife approach (random selection with replacement
of the complete set of instruments) can be used to obtain a sampling
distribution for the model estimate amongst these four variations. Using
1000 rounds of jackknife estimates, we can obtain a final estimate using
the mean or the median of the distribution. We only use the jackknife
approach for associations where there are 15 or more variants in order
to avoid saturating the possible number of instrument combinations.

@@

\textbf{Median-based methods:} An alternative approach is to take the
median effect of all available instruments (4,22). This has the
advantage that up to half the instruments can be invalid, and the
estimate will remain unbiased. Developing the approach further to allow
stronger instruments to contribute more towards the estimate can be
obtained by obtaining the median of the weights of each instrument. The
penalised weighted median estimator introduces a further weight to the
instruments, penalising any instrument that contributes substantially
towards the heterogeneity statistic. Together, this provides three
median based estimators. Other estimation strategies not considered
here, such as LASSO regression, have also been developed for the
situation where at least half of the instruments are valid (23).

\textbf{Mode-based methods:} Similar to the median, the mode based
estimator clusters the instruments into groups based on similarity of
causal effects, and returns the final causal effect estimate based on
the cluster that has the largest number of instruments (6). There are
four implementations of this method: the simple and the weighted mode,
each weighted with or without the assumption of no measurement error in
the exposure estimates (NOME). The simple mode is the unweighted mode of
the empirical density function of causal estimates, whereas the weighted
mode is weighted by the inverse variance of the outcome effect. The
bandwidth parameter was set to 1 by default.

@@

The simplest approach to selecting instruments for performing MR is to
take SNPs that have been declared significant in the published GWAS for
the exposure. This typically involves obtaining SNPs that surpass
\(p < 5 \times 10^{-8}\), using clumping to obtain independent SNPs, and
then replicating in an independent sample. These results are often
recorded in public GWAS catalogs. Alternatively the clumping procedure
can be performed using complete summary data in MR-Base (11). We call
this the ``top hits'' strategy.

@@

With genome-wide association studies growing ever larger, the
statistical power to detect significant associations that may be
influencing the trait downstream of many other pathways increases. For
example, if a SNP \(g_{A}\) influences trait \(A\), and trait \(A\)
influences trait \(B\), then a sufficiently powered GWAS will identify
the \(g_{A}\) as being significant for trait \(B\) (Figure 1a). Using
\(g_{A}\) as an instrument to test the causal effect of \(A\) on \(B\)
is perfectly valid. But in the (incorrectly hypothesised) MR analysis of
trait \(B\) on trait \(A\) could erroneously result in the apparent
causal association of \(B\) on \(A\). If \(g_{A}\) is only one of many
known instruments for \(B\), amongst which some are valid, it is to the
advantage of the researcher to exclude \(g_{A}\) from the analysis.

An approach to inferring the causal direction between phenotypes
recently developed (7) uses the following basic premise. If trait \(A\)
causes trait \(B\) then

@@

because the \(cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}\). This
simple inequality will not hold in some cases, for example
\(\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}\) where \(\rho_{x, x_o}\) and
\(\rho_{y, y_o}\) are the precision of the measurements of the \(x\) and
\(y\). Some combinations of confounding effects can also distort the
\(\rho_{g,x}\) and \(\rho_{g,y}\) parameters, as has been discussed in
detail previously. However, we use it here as a computationally
inexpensive and approximate method to identify variants that are likely
to be invalid. Steiger's Z-test of correlated correlations (24) can be
used to formally test the extent to which the two correlations are
statistically different.

Here we adapt this approach to automatically filter SNPs that are liable
to be invalid (Figure 1a). Other methods have been developed to identify
invalid instruments for the purposes of exclusion from MR analyses
(8,12,25), based on the notion that outlying instruments are likely to
be exhibiting horizontal pleiotropy. A potential drawback of this
approach is that the outlier SNPs in a summary-set might be the only
reliable ones. This could arise, for example, for the \emph{CRP} variant
instrumenting C-reactive protein levels (19), or the \emph{SLC2A9}
variant influencing serum urate levels (26).

@@

We consider the 14 MR methods described above, for which instruments can
be supplied using two instrument selection strategies, leading to 28
methods in total (Supplementary table 2). In the context of this
analysis, each method is considered to be an `expert', taking a set of
SNP-exposure and SNP-outcome effect sizes and their standard errors as
inputs. Our objective is to select the expert most likely to be correct
for a specific MR analysis. An overview of the approach is shown in
Figure 2.

The mixture of experts method is a machine learning approach (27) which
seeks to divide a parameter space into subdomains, such that a
particular expert is used primarily for problems that reside in a
subdomain most suited to that expert. In this case our objective is to
identify characteristics of the SNP-exposure and SNP-outcome
associations for which one specific MR method is most likely to yield
highest statistical power for non-null associations, and lowest false
discovery rates for null associations. This involves creating a `gating
function', whose purpose is to decide which expert to use for a specific
MR analysis, given the parameter space that is occupied by that
summary-set. The metrics are a mixture of regression diagnostics and are
described in Supplementary table 3.

The gating function needs to be trained using data for which the true
causal effect is known, and to this end we generated a large number of
simulated summary-sets. We trained the gating function using random
forest learning algorithms that seek to identify sectors of the
parameter space that are most likely to return accurate estimates for a
particular expert. Figure 2a illustrates how the gating function is
trained using simulated data. New summary-sets can then be applied to
the trained mixture of experts (MoE) model (Figure 2b). We call this
implementation MR-MoE 1.0. Full details are provided in the Online
Methods. MR-MoE 1.0 is implemented in the TwoSampleMR R package
available at
\href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}
(11).

@@

The set of MR estimates obtained from this analysis, and the
summary-sets used to generate them, are deposited in a Neo4j graph
database. Here nodes representations exist for traits and SNPs; and
relationship representations exist for SNP-trait associations and
trait-trait MR estimates. Because each trait-trait association has up to
28 different MR estimates, for simplicity we also distill this down to a
third relationship type comprising a single `best estimate' for each
trait-trait association using the following rules:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  If the number of variants after Steiger filtering is greater than 5
  then apply the MoE to obtain the best method
\item
  If the number of variants after Steiger filtering is less than or
  equal to 5 but greater than 1 then use the IVW random effects approach
  on the filtered set of variants
\item
  If there is 1 variant retained in the Steiger approach then use the
  Wald ratio on the remaining variant
\item
  If there are no variants remaining after Steiger filtering then
  declare no causal association.
\end{enumerate}

@@

As statistical power for GWAS studies improves the likelihood of a
significant association being discovered that doesn't act primarily on
the trait of interest increases (Figure 1a). We evaluated the efficacy
of the Steiger filtering approach for improving instrument selection
using 100,000 simulated summary-sets comprising both null and non-null
causal models. For each summary-set, instruments were selected based on
the tophits strategy and the Steiger filtering strategy, and MR was
performed using 14 different methods based on instruments selected from
each of these strategies.

Figure 1b shows that the tophits strategy led to over half of the
instruments being primarily associated with either confounders or the
outcome phenotype, not the exposure phenotype. The proportion of invalid
instruments due to reverse cause increased as GWAS discovery sample size
increased. Applying Steiger filtering reduced this to 25\%.
Consequently, the false discovery rates (FDR) for 12 of the 14 methods
reduced substantially when applied using Steiger filtered instruments
(Figure 1c). The true positive rates for the methods based on Steiger
filtering did however reduce slightly for 10 of the 14 methods.

@@

We applied our analysis using summary data for 2407 phenotypes,
including 149 complex traits and diseases, 575 metabolites (28,29) and
1683 plasma protein levels (30). For the protein levels only the
instruments were available, so they could only be evaluated as exposure
phenotypes. The complex traits and diseases and metabolite levels had
complete summary data available in MR-Base, thus they be evaluated as
both exposures (if they had significant instruments) and outcomes.
Together, we evaluated 715681 relationships. The majority of these
associations could only be evaluated using fewer than 5 SNPs, and so the
Wald ratio or IVW fixed effects methods were used, but for 61029
associations the MR-MoE approach was applied.

@@

Figure 4: Lookup of causal associations involving a) LDL cholesterol and
b) Years of Schooling. In a) the nodes are filtered to not include those
obtained from the Shin et al or Kettunen et al metabolomic studies. The
arrows denote causal direction and the values on the arrows denote the
causal effect estimate. Only those relationships are shown for which the
\(FDR < 0.05\).

>>>>

\setlength{\droptitle}{-2em}
  \title{Learning pleiotropic signatures to construct a causal map of the human
phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{02 August 2017}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  MRC Integrative Epidemiology Unit (IEU) at the University of Bristol,
  Population Health Sciences, Bristol, UK
\item
  Intelligent Systems Laboratory, Department of Computer Science,
  University of Bristol
\end{enumerate}

@@

A major application for genome-wide association studies (GWAS) has been
the emerging field of causal inference using Mendelian randomization
(MR), where the causal effect between a pair of traits can be estimated
using only summary level data. MR depends on SNPs exhibiting vertical
pleiotropy, where the SNP influences an outcome phenotype only through
an exposure phenotype. Issues arise when this assumption is violated due
to SNPs exhibiting horizontal pleiotropy, and many methods have been
developed in an attempt to address this, each assuming a different model
of horizontal pleiotropy. Here we show that across a range of horizontal
pleiotropy models instrument selection will be increasingly liable to
selecting invalid instruments as GWAS sample sizes continue to grow. We
have integrated these methods into a mixture-of-experts machine learning
approach (MR-MoE 1.0) that predicts the most appropriate model to use
for any specific causal analysis, improving on both power and false
discovery rates. Using the approach, we systematically estimated the
causal effects amongst 2407 phenotypes, generating a working draft of
the causal map of the human phenome, comprising the evaluation of
causality between over a million pairs of traits. Almost 90\% of causal
estimates indicated some level of horizontal pleiotropy. The causal
estimates are organised into a publicly available graph database
(\url{http://eve.mrbase.org}).

@@

Mendelian randomization (MR) (1,2) exploits vertical pleiotropy to infer
the causal relationships between phenotypes. Suppose that one trait (the
exposure) causally influences another (the outcome). If a SNP influences
the outcome through the exposure then the SNP is exhibiting vertical
pleiotropy (see Box 1). Such a genetic variant is considered to be a
valid instrumental variable if it only influences the outcome through
the exposure (the exclusion restriction assumption). Vertical pleiotropy
can be exploited to mimic a randomised controlled trial, enabling a
causal estimate to be made by comparing the outcome phenotypes between
those individuals that have the exposure-increasing allele against those
who do not, although not without caveats (3). Multiple independent
genetic variants for a particular exposure can be used jointly to
improve causal inference, because a) each variant represents an
independent natural experiment, and an overall causal estimate can be
obtained by meta-analysing the single estimates from each instrument;
and b) potential bias arising from the exclusion restriction assumption
can be detected or corrected by evaluating the consistency of effects
across instruments (4--9).

Genome-wide association studies (GWAS) have identified potential
instrumental variables for thousands of phenotypes (10). Recent
developments in MR have enabled knowledge of instrumental variables to
be applied using either individual-level or only summary-level data
(known as two-sample MR, 2SMR) (11). Here, in order to infer the causal
effect of an exposure on an outcome all that is required are the genetic
effects of the instrumenting SNP on the exposure and the outcome. This
has three major advantages. First, GWAS summary data are non-disclosive
and often publicly available. Second, causal inference can be made
between phenotypes even if they have not been measured in the same
samples, limiting possible MR analyses only by the availability of GWAS
summary data for the traits in question (12). Third, because each
instrumental variable mimics an independent randomization study, we can
view the causal inference problem through the simple and widely
understood prism of a meta-analysis.

Problems with obtaining unbiased causal effects can arise, however, if
the genetic instruments exhibit horizontal pleiotropy (HP), where they
influence the outcome through a pathway other than the exposure (2). The
extent of this phenomenon is not to be understated, and many methods
have been developed that attempt to reliably obtain unbiased causal
estimates under specific models of HP (4--7,9,13). It is considered best
practice to report estimates from a range of methods as a sensitivity
analysis when presenting causal estimates so as to assess the potential
for different horizontal pleiotropy models and other violations of MR
assumptions. However it is also desirable to consolidate across methods
under some circumstances. First, it could be of interest to screen links
between thousands of traits for being potential causal associations, in
which case a critical evaluation of each causal estimate of interest may
not be possible or convenient. Second, though the simplest method, based
on inverse-variance weighted (IVW) meta analysis (11), is most
statistically powerful under no HP, it can have high false positive or
low true positive rates in the presence of HP compared to other methods.
Pleiotropy has been hypothesised to be universal (14), though the degree
and nature (e.g.~horizontal versus vertical) to which this may be the
case is contested (15,16), hence defaulting to the IVW method in the
first instance and using other methods as sensitivity analyses may not
be appropriate. Third, if different methods disagree it is not possible
to know which is correct because the true nature of HP exhibited by the
instruments is not known. Fourth, the available methods do not cover all
possible models of HP, and therefore an automated method for instrument
selection may be necessary.

In this paper we introduce two innovations towards improving the
reliability of MR estimates. First, we introduce an approach to discard
genetic variants that are likely to be invalid. Second, we hypothesised
that characteristics of the summary data could indicate which method
would be most reliable, and we introduce new machine learning approaches
that attempt to automate both instrument and method selection. Using
curated GWAS summary data for thousands of phenotypes (12), we use these
new methods to construct a graph of millions of causal estimates.
Motivated by the recent avalanche (17) of 2SMR publications, with
similar (18,19) or contradicting (20,21) conclusions despite using the
same data (17), we developed a graph database to represent these
estimates in a consistent manner. We consider this to be a `working
draft' of the causal map of the human phenome, but raise caution
throughout that its interpretation is far from straightforward, and that
future corrections and refinements will inevitably follow as data grows
and 2SMR methods evolve.

\subsubsection{Box 1: Models of
pleiotropy}\label{box-1-models-of-pleiotropy}

In this paper we separate pleiotropy into two forms for simplicity -
vertical and horizontal. Vertical pleiotropy has also been known as
mediated pleiotropy (22), because the genetic variant influences one
trait, which in turn influences a second trait. It has also been termed
Type II pleiotropy (15) and secondary pleiotropy (23).

By contrast, horizontal pleiotropy can manifest in a myriad of different
ways, but the guiding principle is that the genetic variant influences
two different phenotypes through independent pathways. Horizontal
pleiotropy has also been termed biological pleiotropy (22), Type I
pleiotropy (15), developmental pleiotropy or selectional pleiotropy
(24). From a MR statistical perspective, one can be more specific and
state that horizontal pleiotropy occurs where a SNP influences an
outcome trait through some pathway other than the exposure trait.

For illustration, Hu et al (2016) (Figure 3) enumerate the biological
mechanisms that can lead to pleiotropy: a single variant influences
multiple genes; a single gene generates multiple gene products; a single
gene product has multiple cellular functions; or a single cellular
function influences multiple phenotypes (25). Whether we consider each
of these mechanisms to be vertical pleiotropy (advantageous for MR) or
horizontal pleiotropy (problematic for MR) depends on which phenotypes
are being interrogated. For example, consider a scenario where a SNP
influences the expression levels of two genes, and the first gene
influences trait A while the second gene influences trait B. Then:

\begin{itemize}
\tightlist
\item
  The genetic variant is exhibiting a vertical pleiotropic effect on
  trait A through the first gene expression level. Here, all other
  assumptions being met, the genetic variant would be a valid instrument
  if used to estimate the causal effect of the gene expression level on
  trait A. However,
\item
  the SNP is exhibiting a horizontal pleiotropic effect on trait A and
  trait B through the two gene expression levels. Here the genetic
  variant would be invalid if used to estimate the causal effect of
  trait A on trait B.
\end{itemize}

@@

The use of summary data in two-sample MR is described in detail
elsewhere (12,26). A brief outline of the procedure is as follows.
First, genetic instruments for the exposure trait need to be identified
- those SNPs with \(p < 10^{-8}\) are retained in order to ensure that
the first assumption of MR (that the instrument associates with the
exposure) is generally satisfied. We collect their effect sizes,
standard errors and effect alleles for the association with the exposure
trait. These can be obtained from published GWAS summary data, either
from from curated lists of established GWAS associations (10) or study
websites comprising complete summary data (all SNP effects across the
genome regardless of strength of association). Next, the effects of
those SNPs on the outcome need to be obtained, typically necessitating
access to complete summary data because these SNPs will not usually
reach genome-wide significance for the outcome trait (and are therefore
usually absent in curated catalogues).

We use the term \emph{summary-set} to refer to the minimum data required
to perform 2SMR analysis - four columns comprising the SNP-exposure and
SNP-outcome effects and their standard errors, with each rows
corresponding to a SNP that is used as an instrument for the exposure.

In its simplest implementation, the regression of the SNP-exposure
effect sizes against the SNP-outcome effect sizes, with greater weight
afforded to those SNPs with smaller SNP-outcome standard errors,
provides the estimate of the causal effect of the exposure on the
outcome. This is known as the inverse-variance weighted (IVW) method.

Given summary data for a large number of traits, it is straightforward
to exhaustively analyse the causal relationships of every trait against
every other trait for which there are sufficient summary data available.
Supplementary table 1 provides a list of all traits that have available
GWAS summary data that were used in these analyses.

@@

In this paper we consider three main classes of MR estimation. Full
details for each approach have been described previously. A summary of
the methods is given in Supplementary table 2. Though other methods also
exist we have limited the analyses to the following 14 methods for
simplicity.

\textbf{Mean-based methods:} Here we consider four nested models (6).
The inverse variance weighted (IVW) fixed effects meta-analysis approach
assumes that variants exhibit no HP. IVW random effects meta-analysis
relaxes the HP assumption, allowing it to be present but balanced - such
that it only leads to increased heterogeneity around the regression line
without affecting the slope (and therefore not introducing bias). Fixed
effects Egger regression (4) relaxes the HP assumption further by
allowing a non-zero intercept which essentially allows overall
horizontal pleiotropy to be directional, where its total effect
influences the outcome in a specific direction. Random effects Egger
regression further allows heterogeneity around the slope having
accounted for overall directional HP (6), as long as the HP effects are
not correlated with the SNP-exposure effects (also known as the INSIDE
assumption)(4).

The Rucker framework (27), adapted to MR (6) uses estimates of
heterogeneity in the IVW and Egger frameworks to navigate between these
nested models. A jackknife approach (random selection with replacement
of the complete set of instruments) can be used to obtain a sampling
distribution for the model estimate amongst these four variations. Using
1000 rounds of jackknife estimates, we can obtain a final estimate using
the mean or the median of the distribution. We only use the jackknife
approach for MR analyses where there are 15 or more instruments, in
order to avoid saturating the possible number of instrument
combinations.

@@

\textbf{Median-based methods:} An alternative approach is to take the
median effect of all available instruments (5,28). This has the
advantage that only half the instruments need to be valid, and the
estimate will remain unbiased. The weighted median estimate develops the
approach further to allow stronger instruments to contribute more
towards the estimate can be obtained by weighting the contribution of
each instrument by the inverse of its variance. The penalised weighted
median estimator introduces a further weight to the instruments,
penalising any instrument that contributes substantially towards the
heterogeneity statistic. Together, this provides three median based
estimators. Other estimation strategies not considered here, such as
LASSO regression, have also been developed for the situation where at
least half of the instruments are valid (29).

\textbf{Mode-based methods:} The mode based estimator clusters the
instruments into groups based on similarity of causal effects, and
returns the final causal effect estimate based on the cluster that has
the largest number of instruments (7). There are four implementations of
this method: the simple and the weighted mode, each weighted with or
without the assumption of no measurement error in the exposure estimates
(NOME). The simple mode is the unweighted mode of the empirical density
function of causal estimates, whereas the weighted mode is weighted by
the inverse variance of the outcome effect. The bandwidth parameter was
set to 1 by default.

@@

The simplest approach to selecting instruments for performing MR is to
take SNPs that have been declared significant in the published GWAS for
the exposure. This typically involves obtaining SNPs that surpass
\(p < 5 \times 10^{-8}\), using clumping to obtain independent SNPs, and
then replicating in an independent sample. These results are often
recorded in public GWAS catalogs. Alternatively the clumping procedure
can be performed using complete summary data in MR-Base (12). Complete
summary data refers to association results for all SNPs used in a GWAS
i.e.~not only those passing statistical thresholds for significance. We
call this the ``tophits'' strategy.

@@

With genome-wide association studies growing ever larger, the
statistical power to detect significant associations that may be
influencing the trait downstream of many other pathways increases. For
example, if a SNP \(g_{A}\) influences trait \(A\), and trait \(A\)
influences trait \(B\), then a sufficiently powered GWAS will identify
the \(g_{A}\) as being significant for trait \(B\) (Figure 1a). Using
\(g_{A}\) as an instrument to test the causal effect of \(A\) on \(B\)
is perfectly valid. But the (incorrectly hypothesised) MR analysis of
trait \(B\) on trait \(A\) could erroneously result in the apparent
causal effect of \(B\) on \(A\). If \(g_{A}\) is only one of many known
instruments for \(B\), amongst which some are valid, it is to the
advantage of the researcher to exclude \(g_{A}\) from the analysis.

An approach to inferring the causal direction between phenotypes (8)
uses the following basic premise. If trait \(A\) causes trait \(B\) then

@@

because the \(cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}\). This
simple inequality will not hold in some cases, for example
\(\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}\) where \(\rho_{x, x_o}\) and
\(\rho_{y, y_o}\) are the precision of the measurements of the \(x\) and
\(y\). Some combinations of confounding effects can also distort the
\(\rho_{g,x}\) and \(\rho_{g,y}\) parameters, as has been discussed in
detail previously (8). However, we use it here as a computationally
inexpensive and approximate method to identify variants that are likely
to be invalid. Steiger's Z-test of correlated correlations (30) can be
used to formally test the extent to which the two correlations are
statistically different.

Here we adapt this approach to automatically filter SNPs that are liable
to be invalid (Figure 1a). Other methods have been developed to identify
invalid instruments for the purposes of exclusion from MR analyses
(9,13,31), based on the notion that outlying instruments are more likely
to be a source of HP. A potential drawback of this approach is that the
outlier SNPs in a summary-set might be the only reliable ones. This
could arise, for example, for the \emph{CRP} variant instrumenting
C-reactive protein levels (20), or the \emph{SLC2A9} variant influencing
serum urate levels (32).

@@

We consider the 14 MR methods described above, for which instruments can
be supplied using two instrument selection strategies, leading to 28
methods in total (Supplementary table 2). Each method assumes or
performs best for a different model of pleiotropy. Our objective is to
select the method most likely to be correct for a specific MR analysis
by predicting the model of pleiotropy, and relating that prediction to
the most appropriate model. To achieve this we use a ``mixture of
experts'' machine learning approach (33), where each MR method is
considered to be an `expert', taking a summary-set as its input. An
overview of the approach is shown in Figure 2.

The mixture of experts method seeks to divide a parameter space into
subdomains, such that a particular expert is used primarily for problems
that reside in a subdomain most suited to that expert. In this case we
first identify characteristics of the SNP-exposure and SNP-outcome
associations for which one specific MR method is most likely to yield
highest statistical power for non-null associations, and lowest false
discovery rates for null associations. This involves creating a `gating
function', whose purpose is to decide which expert to use for a specific
MR analysis, given the parameter space that is occupied by that
summary-set. The metrics are a collection of regression diagnostics and
are described in Supplementary table 3.

The gating function needs to be trained using data for which the true
causal effect is known, and to this end we generated a large number of
simulated summary-sets. We trained the gating function using random
forest learning algorithms that seek to identify sectors of the
parameter space that are most likely to return accurate estimates for a
particular expert. Figure 2a illustrates how the gating function is
trained using simulated data. New summary-sets can then be applied to
the trained mixture of experts (MoE) model (Figure 2b). We call this
implementation MR-MoE 1.0. Full details are provided in the Online
Methods. MR-MoE 1.0 is implemented in the TwoSampleMR R package
available at
\href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}
(12).

It should be noted that in the original hierarchical mixture of experts
approach of (33) the gating function and the experts share the same
input space and are trained simultaneously using
expectation-maximisation. In our approach the gating function is defined
over a separate input space consisting of the parameter space of the
experts, and is trained separately in a supervised setting using
simulated data. As such our MR-MoE approach fits the framework of
meta-learning (34--36) where machine learning methods are applied at the
`meta-level' to analyse the results of machine learning experiments at
the `base-level', in order to be able to understand and model the
capabilities of each expert and recommend which expert should be applied
to a given problem. From this perspective the 28 MR methods are the
base-models, the gating function is the meta-model, and its
meta-features are the parameters of the summary-sets.

@@

The set of MR estimates obtained from this analysis, and the
summary-sets used to generate them, are deposited in a Neo4j graph
database. Here node representations exist for traits and SNPs; and
relationship representations exist for SNP-trait associations and
trait-trait MR estimates. Because each trait-trait association has up to
28 different MR estimates, for simplicity we also distill this down to a
third relationship type comprising a single `best estimate' for each
trait-trait association using the following rules:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  If the number of variants after Steiger filtering is greater than 5
  then apply the MoE to obtain the best method. This value is chosen
  arbitrarily as a minimum number of variants for which the different MR
  methods can model HP
\item
  If the number of variants after Steiger filtering is less than or
  equal to 5 but greater than 1 then use the IVW random effects approach
  on the filtered set of variants
\item
  If there is 1 variant retained in the Steiger approach then use the
  Wald ratio on the remaining variant
\item
  If there are no variants remaining after Steiger filtering then
  declare no estimate of a causal association possible.
\end{enumerate}

@@

As statistical power for GWAS studies improves the likelihood of a
significant association being discovered for an outcome trait that acts
primarily on some exposure trait for that outcome increases (Figure 1a).
We evaluated the efficacy of the Steiger filtering approach for
improving instrument selection using 100,000 simulated summary-sets
comprising both null and non-null causal models. For each summary-set,
instruments were selected based on the tophits strategy and the Steiger
filtering strategy, and MR was performed using 14 different methods
based on instruments selected from each of these strategies.

Figure 1b shows that the tophits strategy led to over half of the
instruments being primarily associated with either confounders or the
outcome phenotype, not the exposure phenotype. For brevity, we refer to
the latter type of invalid instruments as ``reverse causal''
instruments. The proportion of invalid instruments due to reverse cause
increased as GWAS discovery sample size increased. Applying Steiger
filtering reduced this to 25\%. Consequently, the false discovery rates
(FDR) for 12 of the 14 methods reduced substantially when applied using
Steiger filtered instruments (Figure 1c). The true positive rates for
the methods based on Steiger filtering did however reduce slightly for
10 of the 14 methods.

@@

We applied our analysis using summary data for 2407 phenotypes,
including 149 complex traits and diseases, 575 metabolites (37,38) and
1683 plasma protein levels (39). For the protein levels only GWAS
significant hits were available, so they could only be evaluated as
exposure phenotypes. The complex traits and diseases and metabolite
levels had complete summary data available obtained via MR-Base, and
thus they could be evaluated as both exposures (if they had significant
instruments) and outcomes. Together, we evaluated 715681 relationships.
The majority of these associations could only be evaluated using fewer
than 5 SNPs, and so the Wald ratio or IVW fixed effects methods were
used, but for 61029 associations the MR-MoE approach was applied.

@@

Figure 4: Lookup of causal associations involving a) LDL cholesterol and
b) Years of Schooling. In a) the nodes are filtered to not include those
obtained from the metabolomic studies (37,38). The arrows denote causal
direction and the values on the arrows denote the causal effect
estimate. Only those relationships are shown for which the
\(FDR < 0.05\).
